The Courtroom Challenge—A Conjecture on the Unavoidable Cost of LLM Verification

### 2025-06-24 paper/main.tex
Add files via upload

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Setup}\label{sec:setup}

\paragraph{Per-round notation.}
Fix a margin parameter $\delta\in(0,1]$.
For every run $\omega$ and round $t=1,\dots,T$ set
\begin{align*}
  y_t      &:=y_t(\theta)          &&\text{(ground-truth label)},\\
  h_t      &:=h_t(\omega)          &&\text{(validator’s automatic prediction)},\\
  \mathrm{Esc}_t&:=\mathbf 1\{\text{round $t$ is escalated}\}.
\end{align*}
Define the \emph{leverage rounds}
\begin{equation}\label{eq:kappaDef}
  \mathcal L(\omega):=\Bigl\{\,t:\,
     \mathrm{Esc}_t=0\;\text{and}\;|y_t-h_t|\ge\delta\Bigr\},
  \qquad
  \kappa(\omega):=\frac{|\mathcal L(\omega)|}{T}.
\end{equation}
The cumulative label error on those rounds is
\begin{equation}\label{eq:ErrDef}
  \Err(\omega):=\sum_{t\in\mathcal L(\omega)}
                \mathbf 1\!\bigl[h_t\neq y_t\bigr].
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Information-theoretic lower bound}\label{sec:info-bound}

\begin{lemma}[Interactive-Fano corollary]\label{lem:ErrLower}
Let $V$ be any validator run for $T$ rounds.
With $\kappa(\omega)$ and $\Err(\omega)$ defined in
\eqref{eq:kappaDef}–\eqref{eq:ErrDef}, write
\[
  \bar\kappa:=\E[\kappa],
  \qquad
  \Err_{\mathrm{bar}}:=\E[\Err].
\]
Assume the transcript $X$ obeys the information budget
\(I(\theta;X)\le I\) under the \emph{uniform} prior on an $M$-point
label set ($M\ge2$).  Then
\begin{equation}\label{eq:ErrLower-new}
  \boxed{\;
    \Err_{\mathrm{bar}}
      \;\ge\;
      \frac{\delta\,T}{2}\;
      \Bigl(\bar\kappa+\gamma-1\Bigr),
      \quad
      \gamma:=\frac12\Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr)
  }.
\end{equation}
In particular, the bound is positive—and linear in $T$—whenever the
validator escalates fewer than a $(1-\gamma)$-fraction of rounds
\((\bar\kappa>1-\gamma)\).
\end{lemma}

\begin{proof}[Sketch]
Step 1 (separation).
For any $m\neq m'$ and transcript $x$ at most one of
$y_t(m),y_t(m')$ can equal $h_t$ on a leverage round, hence
\[
  \Err_m(x)+\Err_{m'}(x)\;\ge\;\delta\,\kappa(\omega)\,T.
  \tag{$\ast$}
\]

\noindent
Step 2 (interactive Fano). 
Set the path-dependent loss
\(L^\star(m,x):=\mathbf 1\!\bigl[\Err_m(x)>
   \tfrac12\delta\kappa(\omega)T\bigr]\)
and $\Delta=\tfrac12$ in Proposition 3 of
\citet{chen2024interfano}.
Inequality ($\ast$) fulfils the separation requirement
\citep[Eq.\,(5)]{chen2024interfano}; thus
\[
  \Pr\!\bigl[\Err(\theta,X)>
     \tfrac12\delta\kappa(\omega)T\bigr]
  \;\ge\;\gamma.
  \tag{$\dagger$}
\]

\medskip\noindent
Step 3 (tail $\Rightarrow$ expectation, expanded).
With
\[
  \mathcal E:=\Bigl\{\Err>\tfrac12\delta\,\kappa T\Bigr\},
  \qquad
  \kappa_{\!*}:=\E[\kappa\mid\mathcal E],
\]
apply the basic fact $\E[Z]\ge a\,\Pr(Z>a)$ to
\(Z=\Err\) and \(a=\tfrac12\delta\,\kappa T\):
\[
\begin{aligned}
  \Err_{\mathrm{bar}}
    &=\E[\Err] \\[2pt]
    &\ge
      \E\!\Bigl[
        \tfrac12\delta\,\kappa T\;
        \mathbf 1_{\{\Err>\tfrac12\delta\,\kappa T\}}
      \Bigr] \\[6pt]
    &=\frac{\delta T}{2}\,
      \E\!\bigl[\kappa\,\mathbf 1_{\mathcal E}\bigr] \\[6pt]
    &=\frac{\delta T}{2}\,
      \kappa_{\!*}\;
      \Pr(\mathcal E).
\end{aligned}
\]

@@

\paragraph{Discussion.}
Write the bound as  
\(\Err_{\mathrm{bar}}\!\ge(\delta T/2)\,\max\{0,\bar\kappa+\gamma-1\}\).
Because \(\gamma>0\) whenever the transcript leaks
fewer than \(\log M-\log2\) bits, any validator that
handles a \emph{non-trivial} share of rounds automatically
\((\bar\kappa>1-\gamma)\)
faces a \emph{linear-in-$T$} unavoidable error floor.
Escalating a larger fraction of rounds (making $\bar\kappa$ smaller) or
relaxing the margin $\delta$ are the only ways to suppress this growth,
precisely matching the “friction buys accuracy’’ narrative.

>>>>

\subsection{Setup}\label{sec:setup}

\paragraph{Per-round notation.}
Fix a margin parameter $\delta\in(0,1]$.
For every run $\omega$ and round $t=1,\dots,T$ set
\begin{align*}
  y_t      &:=y_t(\theta)          &&\text{(ground-truth label)},\\
  h_t      &:=h_t(\omega)          &&\text{(validator’s automatic prediction)},\\
  \mathrm{Esc}_t&:=\mathbf 1_{\{\text{round $t$ is escalated}\}.}
\end{align*}
Define the \emph{leverage rounds}
\begin{equation}\label{eq:kappaDef}
  \mathcal L(\omega):=\Bigl\{\,t:\,
     \mathrm{Esc}_t=0\;\text{and}\;|y_t-h_t|\ge\delta\Bigr\},
  \qquad
  \kappa(\omega):=\frac{|\mathcal L(\omega)|}{T}.
\end{equation}
The cumulative label error on those rounds is
\begin{equation}\label{eq:ErrDef}
  \Err(\omega):=\sum_{t\in\mathcal L(\omega)}
                \mathbf 1_{\!\bigl[h_t\neq y_t\bigr]}.
\end{equation}
Define the global means:
\[
  \bar\kappa:=\E[\kappa],
  \qquad
  \Err_{\mathrm{bar}}:=\E[\Err].
\]

\subsection{Information-theoretic lower bound}\label{sec:info-bound}

\begin{lemma}[Interactive-Fano corollary]\label{lem:ErrLower}
Let $V$ be any validator run for $T$ rounds.
With $\kappa(\omega)$ and $\Err(\omega)$ defined in
\eqref{eq:kappaDef}–\eqref{eq:ErrDef}, write
\[
  \Err_{\mathrm{bar}}:=\E[\Err].
\]
Assume the transcript $X$ obeys the information budget
\(I(\theta;X)\le I\) under the \emph{uniform} prior on an $M$-point
label set ($M\ge2$).  Then
\begin{equation}\label{eq:ErrLower-new}
  \boxed{\;
     \Err_{\mathrm{bar}}
       \;\ge\;
       \frac{\delta\,T}{4}\;
       \kappa_{\!*}\;
       \Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr),
     \quad
     \kappa_{\!*}:=\E[\kappa\mid\Err>\tfrac12\delta\kappa T]
  }
  \, .
\end{equation}
\end{lemma}

\begin{proof}[Proof sketch]
Step 1 (separation).
For any $m\neq m'$ and transcript $x$ at most one of
$y_t(m),y_t(m')$ can equal $h_t$ on a leverage round, hence
\[
  \Err_m(x)+\Err_{m'}(x)\;\ge\;\delta\,\kappa(\omega)\,T.
  \tag{$\ast$}
\]

\noindent
Step 2 (interactive Fano). 
Set the path-dependent loss
\[
  L^\star(m,x)
    := \mathbf{1}_{\{\Err_m(x)>\tfrac12\,\delta\,\kappa(\omega)\,T\}}
\]
and $\Delta=\tfrac12$ in Proposition 3 of
\citep{chen2024interfano}.
Inequality ($\ast$) fulfils the separation requirement
\citep[Eq.\,(5)]{chen2024interfano}.
Define
\[
  \gamma:=\frac12\Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr)\in(0,\tfrac12].
\]
Then
\begin{equation}\tag{$\dagger$}
  \Pr\!\bigl[\Err(\theta,X)>\tfrac12\delta\,\kappa(\omega)T\bigr]
  \;\ge\;\gamma .
\end{equation}

\medskip\noindent
Step 3 (tail $\Rightarrow$ expectation, expanded).
With
\[
  \mathcal E:=\Bigl\{\Err>\tfrac12\delta\,\kappa T\Bigr\},
  \qquad
  \kappa_{\!*}:=\E[\kappa\mid\mathcal E],
\]
we have:
\[
\begin{aligned}
  \Err_{\mathrm{bar}}
    &=\E[\Err] \\[2pt]
    &\ge
      \E\!\Bigl[
        \tfrac12\delta\,\kappa T\;
        \mathbf{1}_{\{\Err>\tfrac12\delta\,\kappa T\}}
      \Bigr] \\[2pt]
    &=\frac{\delta T}{2}\,
      \E\!\bigl[\kappa\,\mathbf{1}_{\mathcal E}\bigr] \\[2pt]
    &=\frac{\delta T}{2}\,
      \kappa_{\!*}\;
      \Pr(\mathcal E) \\[2pt]
    &\ge \frac{\delta T}{4}\,\kappa_{\!*}\, \Bigl(1-\tfrac{I+\log 2}{\log M}\Bigr)\;.
\end{aligned}
\]

@@

\paragraph{Remark (relating $\kappa_{\!*}$ to $\bar\kappa$).}
Write $p_{\mathrm{bad}}:=\Pr(\mathcal E)$ with 
$\mathcal E=\{\Err>\tfrac12\delta\kappa T\}$.  
Since $0\!\le\!\kappa\!\le\!1$ one has
$\bar\kappa
   =p_{\mathrm{bad}}\kappa_{\!*}
    +(1-p_{\mathrm{bad}})\kappa^{\complement}_{\!*}
   \le p_{\mathrm{bad}}\kappa_{\!*}+1-p_{\mathrm{bad}}$.
Hence  
$\kappa_{\!*}\ge\bar\kappa+p_{\mathrm{bad}}-1
            \ge\bar\kappa+\gamma-1$,
where $\gamma$ is the Fano factor.  Therefore
$\kappa_{\!*}\ge\max\{0,\bar\kappa+\gamma-1\}$ as used in the following
discussion.

\paragraph{Discussion.}
Write the bound as  
\(\Err_{\mathrm{bar}}\!\ge(\tfrac12\delta T)\,\max\{0,\bar\kappa+\gamma-1\}\).
Because \(\gamma>0\) whenever the transcript leaks
fewer than \(\log M-\log2\) bits, any validator that
handles a non-trivial share of rounds automatically
\((\bar\kappa>1-\gamma)\)
faces a linear-in-$T$ unavoidable error floor.
Escalating a larger fraction of rounds (making $\bar\kappa$ smaller) or
relaxing the margin $\delta$ are the only ways to suppress this growth,
precisely matching the “friction buys accuracy’’ narrative.
